---
title: 'Supplementary material for UniqTag: Content-derived unique and stable identifiers for gene annotation'
author: 'Shaun Jackman'
output:
  html_document:
    highlight: pygments
---

# Supplementary material

The following supplementary material of the UniqTag paper present the code, shown in Listings A and B, and the data, shown in supplementary Table A, used to generate Figures 1 and 2 of the main manuscript.

```{r install.packages, eval=FALSE, include=FALSE}
install.packages('ggplot2')
install.packages('kable')
install.packages('reshape2')
install.packages('scales')
```

# Load libraries
```{r libraries, message=FALSE}
library(ggplot2)
library(knitr) # for kable
library(reshape2)
library(scales) # for alpha
```

# Read the data
```{r data}
data.orig <- read.delim('UniqTag-supp.tsv',
	colClasses = c(A = 'factor', B = 'factor'))
x <- do.call(rbind, strsplit(as.character(data.orig$Table), '.', fixed = TRUE))
colnames(x) <- c('Data', 'Transform', 'Identifier')
data <- cbind(data.orig, x)
rm(x)
data$k <- as.integer(gsub('^[a-z]*', '', data$Identifier))

build.wide <- with(data,
	data.frame(Build.A = A, Build.B = B,
		Num.A = Only.A + Both, Num.B = Only.B + Both))
build.tall <- melt(build.wide, id.vars = c('Build.A', 'Build.B'),
	variable.name = 'Build', value.name = 'Count')
```

# Figure 1. Plot the number of common identifiers vs. other build
The number of common UniqTag identifiers between build 75 of the Ensembl human genome and nine other builds, the number of common gene and protein identifiers between builds, and the number of genes with peptide sequences that are identical between builds.
```{r ensembl, fig.width=6, fig.height=5, dpi=150}
data.subset <- subset(data, data$k == 9 | is.na(data$k))
aes.data <- aes(x = A, y = Both,
	group = Table, colour = Identifier)
aes.build <- aes(x = Build.A, y = Count,
	group = Build, linetype = Build, shape = Build)
ggplot() +
	geom_point(aes.data, data.subset) +
	geom_line(aes.data, data.subset) +
	scale_colour_brewer(palette = 'Set1',
		breaks = c('gene', 'uniqtag9', 'id', 'seq'),
		labels = c('Gene ID (ENSG)', 'UniqTag (9-mer)',
			'Protein ID (ENSP)', 'Identical peptide sequence')) +

	geom_point(aes.build, build.tall) +
	geom_line(aes.build, build.tall) +
	scale_linetype_manual(name = 'Number of genes',
		breaks = c('Num.B', 'Num.A'),
		labels = c('Ensembl build 75', 'Other Ensembl build'),
		values = c('solid', 'dashed')) +
	scale_shape_manual(name = 'Number of genes',
		breaks = c('Num.B', 'Num.A'),
		labels = c('Ensembl build 75', 'Other Ensembl build'),
		values = c(20, 32)) +

	theme_bw() +
	theme(legend.position = c(1.0, 0),
		legend.justification = c(1, 0),
		legend.box.just = 'right',
		legend.background = element_rect(fill = alpha('white', 0))) +
	xlab('Other Ensembl build') +
	ylab('Identifiers in common with Ensembl build 75')
```

# Figure 2. Plot the number of common identifiers vs. *k*
The number of common UniqTag identifiers between build 75 of the Ensembl human genome and nine other builds for different values of *k*.
```{r k, fig.width=6, fig.height=5, dpi=150}
ggplot(na.omit(data), aes(x = k, y = Both, group = A, colour = A)) +
	geom_point() +
	geom_line() +
	scale_x_continuous(trans = log_trans(),
		breaks = c(1, 2, 5, 10, 20, 50, 100, 200)) +
	scale_colour_brewer(name = 'Other Ensembl build', palette = 'Set1') +
	guides(colour = guide_legend(reverse = TRUE)) +
	theme_bw() +
	xlab('Size of UniqTag k-mer (aa)') +
	ylab('Identifiers in common with Ensembl build 75')
```

# Listing A. UniqTag 1.0
This listing shows the source of [UniqTag 1.0](data/uniqtag), implemented in Ruby.
```ruby
`r paste(readLines('data/uniqtag'), collapse = '\n')`
```

# Listing B. Calculate the number of common identifiers
This [Makefile](data/Makefile) script calculates the data used to plot the above figures.
```makefile
`r paste(readLines('data/Makefile'), collapse = '\n')`
```

# Table A. The number of common identifiers
These data are used to plot the above figures. They are also available in [tab-separated values (TSV) format](UniqTag-supp.tsv).
```{r data-table, results='asis'}
kable(data)
```
